Copulas and their potential for ecology

### 2019-10-04 Paper.Rmd
done editing Q2 sections, just one thing for Shyamolina to do

<!--***DAN: stopped here for methods of Q2-->

<!--***DAN: How much of the material from Appendix S8 should be moved into
the main text? I am thinking a fair amount of it. Currrently there
is really no results on that topic at all in the main text, and that seems
a bit ridiculous and probably is why the referee said we treated this 
material in a "breezy" way.-->
The next mechanism we investigated is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence between the 
two character values across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
independent draws from bivariate distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For 
each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

<!--***DAN: Decided to remove this simplified version of the figure for AER
and replace by the bigger one which was in the supp mat, now below.
\begin{figure}[!h]
\begin{center}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Example results showing similarity of copula statistics for 
environmental-noise inputs and ecological-variable outputs of a dynamical model.
Asymmetric tail dependence was similar in both noise inputs and model outputs
for this model. The 
AR(1) model and a Clayton copula were used (see methods section for Q2).
Each point is the mean across 50 replicate simulations for which the same 
Kendall $\tau$ value was used. Error bars are standard errors 
and panel headers give Pearson correlation results for
the points. The regression line through the points (black line) was similar
to the 1-1 line (green line). 
See Fig. \ref{SM-fig_Cause4copula_C_kend_ar1} for additional results for the AR(1) model and Clayton copula, 
Figs \ref{SM-fig_Cause4copula_SC_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1} for AR(1) results with other copula families, and 
Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1} for
the nonlinear models described in the methods section for Q2.\label{fig_Cause4copula_scatter_C_kend_ar1}}
\end{center}
\end{figure}
-->

<!--Plots for AR(1) with Clayton-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=13 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/common_legend_cause4copula_stat.pdf}\\
\vspace{-0.4 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Spearman_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Kendall_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pearson_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Coru_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl-Coru_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pu_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl-Pu_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2l_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u-D2l_vs_Kendall's Tau.pdf} \\
\hspace{-0.6 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Comparison of correlation and tail association statistics between 
environmental-noise inputs and ecological-variable 
outputs, AR(1) model, Clayton copula. Red and blue points give 
means over 50 replicate simulations of the listed statistics, 
red points are for noise inputs and blue points are for
model outputs. Error bars are standard errors. 
$P$-values (purple points, right axis) are for a 
paired $t$-test of the null hypothesis that the distributions have the same mean. 
For bottom panels, headers give Pearson correlations for
the points. The regression line through the points (black) was similar
to the 1-1 line (green). 
See methods section for Q2 
for details.\label{fig_Cause4copula_C_kend_ar1}}
\end{center}
\end{figure}
<!--***DAN: Shyamolina, the panels in the bottom row do not line up with the others. 
I remember you fixed this before for the sup mat figs, can you plase fix it again here?
Also, can you make the panels smaller or more tightly packed or soemthing
so the whole figure plus caption fits on a page?
***Shya: Ok, now I fixed it. See the margin specifications are different for Paper and Suuppmat, 
that's why I have to downsize the figure here in the maintext.-->

<!--***DAN: Got to here in results for Q2-->

For our evolutionary model, the hypothesis was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence between characters across
phylogeny tips (Appendix \ref{SM-evol_mech}, Fig. \ref{SM-fig_cop_evo} 
for details). 
Thus this is a mechanism that may produce non-normal copula structure and
asymmetric tail dependence between characters.
This result is represented in Fig. \ref{MasterFigure} as the dashed box around 
"If character evolution has copula structure" and the arrow labelled "D".
This topic is revisited in the Discussion, as is the box "Asymmetric species
interactions" and the arrow labelled "C" in Fig. \ref{MasterFigure}.

>>>>

The next mechanism we investigated is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the bias toward right tail association
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that asymmetric tail association occurs in evolutionary changes
in bivariate characters, and gives rise to asymmtric tail association between the 
two character values across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
independent draws from bivariate distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For 
each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed using our nonparametric
statistics.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
associations in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

<!--***DAN: Decided to remove this simplified version of the figure for AER
and replace by the bigger one which was in the supp mat, now below. Keeping the
code the simpler one here in case we ever need it.
\begin{figure}[!h]
\begin{center}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=5 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Example results showing similarity of copula statistics for 
environmental-noise inputs and ecological-variable outputs of a dynamical model.
Asymmetric tail dependence was similar in both noise inputs and model outputs
for this model. The 
AR(1) model and a Clayton copula were used (see methods section for Q2).
Each point is the mean across 50 replicate simulations for which the same 
Kendall $\tau$ value was used. Error bars are standard errors 
and panel headers give Pearson correlation results for
the points. The regression line through the points (black line) was similar
to the 1-1 line (green line). 
See Fig. \ref{SM-fig_Cause4copula_C_kend_ar1} for additional results for the AR(1) model and Clayton copula, 
Figs \ref{SM-fig_Cause4copula_SC_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1} for AR(1) results with other copula families, and 
Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1} for
the nonlinear models described in the methods section for Q2.\label{fig_Cause4copula_scatter_C_kend_ar1}}
\end{center}
\end{figure}
-->

<!--Plots for AR(1) with Clayton-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=13 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/common_legend_cause4copula_stat.pdf}\\
\vspace{-0.4 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Spearman_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Kendall_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pearson_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Coru_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Corl-Coru_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pu_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_Pl-Pu_vs_Kendall's Tau.pdf}\\
\vspace{-0.2 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2l_vs_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_D2u-D2l_vs_Kendall's Tau.pdf} \\
\hspace{-0.6 cm}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_CorlmCoru_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_PlmPu_Kendall's Tau.pdf}
\includegraphics[width=4.3 cm]{./Results/Cause4copula_results/Cause4copula_stat_results_ar1_beta_0.5_Clayton/Clayton_scatter_D2umD2l_Kendall's Tau.pdf}
\caption{Comparison of correlation and tail association statistics between 
environmental-noise inputs and ecological-variable 
outputs, AR(1) model, Clayton copula. Red and blue points give 
means over 50 replicate simulations of the listed statistics, 
red points are for noise inputs and blue points are for
model outputs. Error bars are standard errors. 
$P$-values (purple points, right axis) are for a 
paired $t$-test of the null hypothesis that the distributions have the same mean. 
For bottom panels, headers give Pearson correlations for
the points. The regression line through the points (black) was similar
to the 1-1 line (green). 
See methods section for Q2 
for details.\label{fig_Cause4copula_C_kend_ar1}}
\end{center}
\end{figure}

For our evolutionary model, the hypothesis was 
correct that asymmetric tail association in evolutionary changes 
can produce similarly asymmetric tail association between characters across
phylogeny tips.
Once character evolution was simulated 100 times for each of the
dependence structure we used on evolutionary changes, 
we had 817 bivariate characters for each of 500 simulations (see methods 
section for Q2). We 
computed our nonparametric asymmetry statistics 
$\cor_{0,0.2}-\cor_{0.8,1}$, $\Ps_{0,0.2}-\Ps_{0.8,1}$,
and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ for each simulation output and 
produced a histogram for each statistic and for each
dependence structure (Fig. \ref{fig_cop_evo}). Results showed that 
asymmetries of tail associations in evolutionary changes
were associated with similar asymmetries of tail associations in extant characters.
Thus, based on our simulations, we cannot reject the possibility that
this is a mechanism behind observed asymmetric tail associations
in our bird and mammal data.
This result is represented in Fig. \ref{MasterFigure} as the dashed box around 
"If character evolution has copula structure" and the arrow labelled "D".
This topic is revisited in the Discussion, as is the box "Asymmetric species
interactions" and the arrow labelled "C" in Fig. \ref{MasterFigure}.

<!--results for third mechanism-->
\begin{figure}[!h]
\textbf{ \hspace{3.5 cm} (A) \hspace{2 cm} (F) \hspace{2 cm} (K)} \\
\vspace{-1 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeL_hist_taildep.pdf}\\
\end{center}
\vspace{-.33 cm}
\textbf{ \hspace{3.5 cm} (B) \hspace{2 cm} (G) \hspace{2 cm} (L)} \\
\vspace{-1 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatL_hist_taildep.pdf}\\
\end{center}
\vspace{-.33 cm}
\textbf{ \hspace{3.5 cm} (C) \hspace{2 cm} (H) \hspace{2 cm} (M)} \\
\vspace{-1 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/BivNormal_hist_taildep.pdf}\\
\end{center}
\vspace{-.33 cm}
\textbf{ \hspace{3.5 cm} (D) \hspace{2 cm} (I) \hspace{2 cm} (N)} \\
\vspace{-1 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatR_hist_taildep.pdf}\\
\end{center}
\vspace{-.33 cm}
\textbf{ \hspace{3.5 cm} (E) \hspace{2 cm} (J) \hspace{2 cm} (O)} \\
\vspace{-1 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeR_hist_taildep.pdf}\\
\caption[Non-parametric statistics on a mechanism of tail dependence in character distributions]{Three 
measures of asymmetry of tail associations ($\cor_{0,0.2}-\cor_{0.8,1}$, A-E; 
$\Ps_{0,0.2}-\Ps_{0.8,1}$, F-J; $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ K-O) between two characters, across extant species,  were 
computed for each of 100 simulations of mammalian character evolution for each of five types of tail dependence between 
evolutionary changes in the characters (extreme left-tail dependence (A, F, K), moderate left-tail dependence (B, G, L), 
symmetry of tail dependence (C, H, M), moderate right-tail dependence (D, I, N), and extreme right-tail dependence 
(E, J, O)). See the methods section for Q2 for details. The number above each panel indicates how many statistics out of $100$ were less than or equal to $0$. Values substantially lower (respectively, higher) than $50$ indicate greater left (respectively, right) tail-dependence between characters.\label{fig_cop_evo}}
\end{center}
\end{figure}
<!--***DAN: Shyamolina, can you please make this figure and its caption fit on one page,
somehow?-->
